---
title: "Gapminder World Dashboard"
format:
  dashboard:
    orientation: rows
    scrolling: true
    theme:
      - flatly
    nav-buttons:
      - text: "Life Expectancy & GDP per Capita"
        href: "#"
    mainfont: "DM Sans"
    include-in-header:
      text: |
        <link href="https://fonts.googleapis.com/css2?family=DM+Sans:wght@300;400;500;600&family=Fraunces:wght@300;400;500&display=swap" rel="stylesheet">
        <style>
          h1, h2, h3, .navbar-brand { font-family: 'Fraunces', serif !important; }
          body, .card-body, .value-box { font-family: 'DM Sans', sans-serif !important; }
          .navbar { background-color: #007BFF !important; }
          .navbar-brand, .navbar-nav .nav-link { color: #ffffff !important; }
        </style>
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
# Load packages
if (!require(pacman)) install.packages("pacman")
pacman::p_load(ggplot2, knitr, tidyr, plotly, janitor, DT, reactable, flexdashboard, gapminder, dplyr, scales, leaflet, sf, rnaturalearth, rnaturalearthdata, bslib, bsicons, htmltools)


# colour palette 
region_colors <- c(
  Africa   = "#2d6a4f",
  Americas = "#1d3461",
  Asia     = "#c94f2a",
  Europe   = "#7b4f9e",
  Oceania  = "#c77d1a"
)

# data preparation 
gap <- gapminder::gapminder %>%
  rename(region = continent) %>%
  mutate(gdp_total = pop * gdpPercap)

gap_2007 <- gap %>% filter(year == 2007)
gap_1952 <- gap %>% filter(year == 1952)

# weighted averages by region × year
region_year <- gap %>%
  group_by(region, year) %>%
  summarise(
    avg_lifeExp  = weighted.mean(lifeExp,  pop, na.rm = TRUE),
    avg_gdpPercap = weighted.mean(gdpPercap, pop, na.rm = TRUE),
    total_pop    = sum(pop, na.rm = TRUE),
    .groups = "drop"
  )

# region summaries for 2007
region_2007 <- region_year %>% filter(year == 2007)

# value-box stats
global_avg_2007   <- weighted.mean(gap_2007$lifeExp, gap_2007$pop)
global_avg_1952   <- weighted.mean(gap_1952$lifeExp, gap_1952$pop)
improvement       <- global_avg_2007 - global_avg_1952
top_country_life  <- gap_2007 %>% slice_max(lifeExp, n = 1)
bot_country_life  <- gap_2007 %>% slice_min(lifeExp, n = 1)
top_country_gdp   <- gap_2007 %>% slice_max(gdpPercap, n = 1)
bot_country_gdp   <- gap_2007 %>% slice_min(gdpPercap, n = 1)
ratio_richpoor    <- round(top_country_gdp$gdpPercap / bot_country_gdp$gdpPercap)

# growth table
growth <- gap %>%
  filter(year %in% c(1952, 2007),
         country %in% c("Japan","South Korea","Norway","Germany",
                         "United States","Brazil","China","India")) %>%
  select(country, year, gdpPercap) %>%
  pivot_wider(names_from = year, values_from = gdpPercap,
              names_prefix = "yr_") %>%
  mutate(gain = yr_2007 - yr_1952)

# outlier table
outliers <- gap_2007 %>%
  filter(country %in% c("Cuba","Kuwait","Gabon","Sri Lanka",
                         "Equatorial Guinea","Japan","Botswana")) %>%
  select(country, region, gdpPercap, lifeExp) %>%
  mutate(Story = case_when(
    country == "Cuba"              ~ "High life exp despite modest GDP universal healthcare",
    country == "Kuwait"            ~ "Oil wealth; good outcomes but modest relative to income",
    country == "Gabon"             ~ "Oil wealth not translating to population health",
    country == "Sri Lanka"         ~ "Strong public health investment despite low income",
    country == "Equatorial Guinea" ~ "Oil boom: very high GDP but poor health outcomes",
    country == "Japan"             ~ "Highest life exp-
    diet, healthcare & social cohesion",
    country == "Botswana"          ~ "HIV/AIDS epidemic depressed life exp despite growth"
  ))
```

# Life Expectancy

## Row {height="160px"}

### Global Average Life Expectancy (2007)

```{r}
bslib::value_box(
  title   = "Global Average Life Expectancy (2007)",
  value   = paste0(round(global_avg_2007, 1), " yrs"),
  showcase = bsicons::bs_icon("globe"),
  theme   = bslib::value_box_theme(bg = "#1d3461", fg = "#ffffff")
)
```

### Highest Life Expectancy (2007)

```{r}
bslib::value_box(
  title   = paste0("Highest: ", top_country_life$country),
  value   = paste0(round(top_country_life$lifeExp, 1), " yrs"),
  showcase = bsicons::bs_icon("arrow-up-circle-fill"),
  theme   = bslib::value_box_theme(bg = "#2d6a4f", fg = "#ffffff")
)
```

### Lowest Life Expectancy (2007)

```{r}
bslib::value_box(
  title   = paste0("Lowest: ", bot_country_life$country),
  value   = paste0(round(bot_country_life$lifeExp, 1), " yrs"),
  showcase = bsicons::bs_icon("arrow-down-circle-fill"),
  theme   = bslib::value_box_theme(bg = "#c94f2a", fg = "#ffffff")
)
```

### Global Improvement Since 1952

```{r}
bslib::value_box(
  title   = "Global improvement since 1952",
  value   = paste0("+", round(improvement, 1), " yrs"),
  showcase = bsicons::bs_icon("graph-up-arrow"),
  theme   = bslib::value_box_theme(bg = "#7b4f9e", fg = "#ffffff")
)
```

## Row {height="420px"}

### Life Expectancy Over Time by Region {width="650"}

```{r title = "Life Expectancy Over Time by Region"}
#| fig-height: 5
plot_ly(region_year,
        x = ~year,
        y = ~avg_lifeExp,
        color = ~region,
        colors = region_colors,
        type = "scatter",
        mode = "lines+markers",
        text = ~paste0("<b>", region, " — ", year, "</b><br>",
                       "Avg Life Exp: ", round(avg_lifeExp, 1), " yrs"),
        hoverinfo = "text") %>%
  layout(
    title = list(text = "Regional trends in life expectancy (1952–2007)",
                 font = list(size = 13)),
    xaxis = list(title = "", tickvals = seq(1952, 2007, 5)),
    yaxis = list(title = "Life Expectancy (years)"),
    legend = list(orientation = "h", y = -0.15)
  )
```

### Average Life Expectancy by Region (2007) {width="650"}

```{r title = "Average Life Expectancy by Region (2007)"}
#| fig-height: 5
region_2007 %>%
  mutate(region = reorder(region, avg_lifeExp)) %>%
  plot_ly(
    x         = ~avg_lifeExp,
    y         = ~region,
    type      = "bar",
    orientation = "h",
    color     = ~region,
    colors    = region_colors,
    text      = ~paste0("<b>", region, "</b><br>", round(avg_lifeExp, 1), " years"),
    hoverinfo = "text",
    textposition = "outside",
    texttemplate = "%{x:.1f}",
    showlegend = FALSE
  ) %>%
  layout(
    title  = list(text = "Life expectancy by region, 2007",
                  font = list(size = 13)),
    xaxis  = list(title = "Life Expectancy (years)",
                  range = c(0, 90),
                  zeroline = FALSE),
    yaxis  = list(title = ""),
    margin = list(l = 10, r = 40),
    bargap = 0.35
  )
```

## Row {height="420px"}

### Top 15 Countries — Life Expectancy (2007) {width="650"}

```{r title = "Top 15 Countries: Life Expectancy (2007)"}
#| fig-height: 5
top15 <- gap_2007 %>%
  slice_max(lifeExp, n = 15) %>%
  mutate(country = reorder(country, lifeExp))

top15 %>%
  plot_ly(
    x            = ~lifeExp,
    y            = ~country,
    type         = "bar",
    orientation  = "h",
    color        = ~region,
    colors       = region_colors,
    text         = ~paste0("<b>", country, "</b> (", region, ")<br>",
                           round(lifeExp, 1), " years"),
    hoverinfo    = "text",
    texttemplate = "%{x:.1f}",
    textposition = "outside",
    showlegend   = FALSE
  ) %>%
  layout(
    title  = list(text = "Top 15 countries, 2007",
                  font = list(size = 12)),
    xaxis  = list(title = "Life Expectancy (years)",
                  range = c(75, 87),
                  zeroline = FALSE),
    yaxis  = list(title = ""),
    margin = list(l = 10, r = 40),
    bargap = 0.3
  )
```

### Bottom 15 Countries — Life Expectancy (2007) {width="650"}

```{r title = "Bottom 15 Countries: Life Expectancy (2007)"}
#| fig-height: 5
bot15 <- gap_2007 %>%
  slice_min(lifeExp, n = 15) %>%
  mutate(country = reorder(country, lifeExp))

bot15 %>%
  plot_ly(
    x            = ~lifeExp,
    y            = ~country,
    type         = "bar",
    orientation  = "h",
    color        = ~region,
    colors       = region_colors,
    text         = ~paste0("<b>", country, "</b> (", region, ")<br>",
                           round(lifeExp, 1), " years"),
    hoverinfo    = "text",
    texttemplate = "%{x:.1f}",
    textposition = "outside",
    showlegend   = FALSE
  ) %>%
  layout(
    title  = list(text = "Bottom 15 countries, 2007",
                  font = list(size = 12)),
    xaxis  = list(title = "Life Expectancy (years)",
                  range = c(35, 60),
                  zeroline = FALSE),
    yaxis  = list(title = ""),
    margin = list(l = 10, r = 40),
    bargap = 0.3
  )
```

## Row {height="520px"}

### World Map — Life Expectancy (2007)

```{r title = "World Map: Life Expectancy (2007)"}
#| fig-height: 5
world_sf <- ne_countries(scale = "medium", returnclass = "sf")

world_joined <- world_sf %>%
  left_join(
    gap_2007 %>% select(country, lifeExp, region),
    by = c("name_long" = "country")
  )

pal <- colorNumeric(
  palette = c("#f5c07a", "#e8a64e", "#c2d48c", "#74c185", "#3a9b6d", "#1a6b4e", "#0a4530"),
  domain  = c(35, 85),
  na.color = "#dddddd"
)

leaflet(world_joined, options = leafletOptions(minZoom = 1)) %>%
  addTiles(options = tileOptions(opacity = 0)) %>%
  addProviderTiles(providers$CartoDB.PositronNoLabels) %>%
  addPolygons(
    fillColor   = ~pal(lifeExp),
    fillOpacity = 0.85,
    color       = "white",
    weight      = 0.5,
    label       = ~paste0(name_long, ": ", ifelse(is.na(lifeExp), "N/A", paste0(round(lifeExp, 1), " yrs"))),
    labelOptions = labelOptions(style = list("font-size" = "13px"))
  ) %>%
  addLegend(
    pal      = pal,
    values   = ~lifeExp,
    title    = "Life Expectancy (yrs)",
    position = "bottomright",
    na.label = "No data"
  ) %>%
  setView(lng = 10, lat = 20, zoom = 2)
```

## Row {height="500px"}

### Country Ranking Table — Life Expectancy (2007) 

```{r title = "Country Ranking Table: Life Expectancy (2007)"}
rank_table <- gap_2007 %>%
  arrange(desc(lifeExp)) %>%
  mutate(
    Rank   = row_number(),
    `Life Exp (yrs)` = round(lifeExp, 1),
    `GDP/cap ($)`    = dollar(round(gdpPercap)),
    Population       = comma(pop)
  ) %>%
  select(Rank, Country = country, Region = region,
         `Life Exp (yrs)`, `GDP/cap ($)`, Population)

DT::datatable(
  rank_table,
  filter    = "top",
  rownames  = FALSE,
  options   = list(
    pageLength = 15,
    dom        = "ftip",
    columnDefs = list(list(className = "dt-center", targets = c(0, 3)))
  ),
  class = "compact stripe hover"
) %>%
  formatStyle(
    "Region",
    backgroundColor = styleEqual(
      names(region_colors), unname(region_colors)
    ),
    color = "white",
    fontWeight = "bold"
  ) %>%
  formatStyle(
    "Life Exp (yrs)",
    background = styleColorBar(range(gap_2007$lifeExp), "#2d6a4f"),
    backgroundSize = "100% 80%",
    backgroundRepeat = "no-repeat",
    backgroundPosition = "center"
  )
```

# GDP per Capita

## Row {height="160px"}

### Global Average GDP/cap (2007)

```{r}
bslib::value_box(
  title   = "Global pop-weighted GDP per capita (2007)",
  value   = dollar(round(weighted.mean(gap_2007$gdpPercap, gap_2007$pop))),
  showcase = bsicons::bs_icon("coin"),
  theme   = bslib::value_box_theme(bg = "#1d3461", fg = "#ffffff")
)
```

### Highest GDP per Capita (2007)

```{r}
bslib::value_box(
  title   = paste0("Highest: ", top_country_gdp$country),
  value   = dollar(round(top_country_gdp$gdpPercap)),
  showcase = bsicons::bs_icon("arrow-up-circle-fill"),
  theme   = bslib::value_box_theme(bg = "#2d6a4f", fg = "#ffffff")
)
```

### Lowest GDP per Capita (2007)

```{r}
bslib::value_box(
  title   = paste0("Lowest: ", bot_country_gdp$country),
  value   = dollar(round(bot_country_gdp$gdpPercap)),
  showcase = bsicons::bs_icon("arrow-down-circle-fill"),
  theme   = bslib::value_box_theme(bg = "#c94f2a", fg = "#ffffff")
)
```

### Richest-to-Poorest Ratio (2007)

```{r}
bslib::value_box(
  title   = "Wealthiest vs. poorest country GDP/cap ratio",
  value   = paste0(ratio_richpoor, "×"),
  showcase = bsicons::bs_icon("bar-chart-fill"),
  theme   = bslib::value_box_theme(bg = "#7b4f9e", fg = "#ffffff")
)
```

## Row {height="420px"}

### GDP per Capita Trends by Region (1952–2007) {width="650"}

```{r title = "GDP per Capita Trends by Region (1952–2007)"}
#| fig-height: 5
plot_ly(
  region_year,
  x         = ~year,
  y         = ~avg_gdpPercap,
  color     = ~region,
  colors    = region_colors,
  type      = "scatter",
  mode      = "lines+markers",
  text      = ~paste0(
                "<b>", region, " — ", year, "</b><br>",
                "Avg GDP/cap: $", comma(round(avg_gdpPercap))
              ),
  hoverinfo = "text"
) %>%
  layout(
    title  = list(text = "Regional GDP per capita trends (1952–2007)",
                  font = list(size = 13)),
    xaxis  = list(title = "",
                  tickvals = seq(1952, 2007, 5),
                  ticktext = seq(1952, 2007, 5)),
    yaxis  = list(title = "GDP per Capita (USD)",
                  tickformat = "$.1fk"),
    legend = list(orientation = "h", y = -0.15),
    margin = list(b = 60)
  )


```

### GDP per Capita by Region — 2007 {width="650"}

```{r title = "GDP per Capita by Region (2007)"}

#| fig-height: 5
region_2007 %>%
  mutate(region = reorder(region, avg_gdpPercap)) %>%
  plot_ly(
    x            = ~avg_gdpPercap,
    y            = ~region,
    type         = "bar",
    orientation  = "h",
    color        = ~region,
    colors       = region_colors,
    text         = ~paste0("<b>", region, "</b><br>$", comma(round(avg_gdpPercap))),
    hoverinfo    = "text",
    texttemplate = ~paste0("$", round(avg_gdpPercap/1000, 1), "k"),
    textposition = "outside",
    showlegend   = FALSE
  ) %>%
  layout(
    title  = list(text = "GDP/cap by region, 2007", font = list(size = 13)),
    xaxis  = list(title = "GDP per Capita (USD)",
                  tickformat = "$.1fk",
                  zeroline = FALSE),
    yaxis  = list(title = ""),
    margin = list(l = 10, r = 60),
    bargap = 0.35
  )
```

## Row {height="420px"}

### Distribution of Countries by Income Bracket (2007) {width="650"}

```{r title = "Distribution of Countries by Income Bracket (2007)"}
#| fig-height: 5
gap_2007_hist <- gap_2007 %>%
  mutate(income_bracket = cut(
    gdpPercap,
    breaks = c(0, 1000, 2000, 5000, 10000, 20000, 40000, Inf),
    labels = c("<$1k","$1k–2k","$2k–5k","$5k–10k","$10k–20k","$20k–40k",">$40k"),
    right  = FALSE
  )) %>%
  count(income_bracket, region) %>%
  group_by(income_bracket) %>%
  mutate(total = sum(n))

plot_ly(
  gap_2007_hist,
  x         = ~income_bracket,
  y         = ~n,
  color     = ~region,
  colors    = region_colors,
  type      = "bar",
  text      = ~paste0(region, ": ", n, " countries"),
  hoverinfo = "text"
) %>%
  layout(
    barmode = "stack",
    title   = list(text = "Countries by income bracket, 2007", font = list(size = 13)),
    xaxis   = list(title = "Income Bracket", tickangle = 30),
    yaxis   = list(title = "Number of Countries"),
    legend  = list(orientation = "h", y = -0.25),
    margin  = list(b = 80)
  )
```

### Top 15 Countries — GDP per Capita (2007) {width="650"}

```{r title = "Top 15 Countries — GDP per Capita (2007)"}
#| fig-height: 5
top15_gdp <- gap_2007 %>%
  slice_max(gdpPercap, n = 15) %>%
  mutate(country = reorder(country, gdpPercap))

top15_gdp %>%
  plot_ly(
    x            = ~gdpPercap,
    y            = ~country,
    type         = "bar",
    orientation  = "h",
    color        = ~region,
    colors       = region_colors,
    text         = ~paste0("<b>", country, "</b> (", region, ")<br>$",
                           comma(round(gdpPercap))),
    hoverinfo    = "text",
    texttemplate = ~paste0("$", round(gdpPercap/1000, 1), "k"),
    textposition = "outside",
    showlegend   = FALSE
  ) %>%
  layout(
    title  = list(text = "Top 15 by GDP/cap, 2007", font = list(size = 12)),
    xaxis  = list(title = "GDP per Capita (USD)",
                  tickformat = "$.1fk",
                  zeroline = FALSE),
    yaxis  = list(title = ""),
    margin = list(l = 10, r = 60),
    bargap = 0.3
  )
```

## Row {height="400px"}

### GDP Growth: 1952 → 2007 for Major Economies

```{r title = "GDP Growth: 1952 - 2007 for Major Economies"}
#| fig-height: 5
growth_long <- growth %>%
  pivot_longer(cols = c(yr_1952, yr_2007),
               names_to  = "year_label",
               values_to = "gdpPercap") %>%
  mutate(year_label = recode(year_label, yr_1952 = "1952", yr_2007 = "2007"),
         country    = reorder(country, -gdpPercap))

plot_ly(
  growth_long,
  x         = ~country,
  y         = ~gdpPercap,
  color     = ~year_label,
  colors    = c("1952" = "#9fb8d8", "2007" = "#1d3461"),
  type      = "bar",
  text      = ~paste0("<b>", country, "</b> (", year_label, ")<br>$",
                      comma(round(gdpPercap))),
  hoverinfo = "text"
) %>%
  layout(
    barmode = "group",
    title   = list(text = "GDP per capita growth 1952 vs 2007, major economies",
                   font = list(size = 13)),
    xaxis   = list(title = "", tickangle = 20),
    yaxis   = list(title = "GDP per Capita (USD)", tickformat = "$.1fk"),
    legend  = list(orientation = "h", y = 1.08),
    margin  = list(b = 60)
  )
```

# Relationship

## Row {height="195px"}

### Countries Analyzed

```{r}
bslib::value_box(
  title   = "Countries in 2007 dataset",
  value   = n_distinct(gap_2007$country),
  showcase = bsicons::bs_icon("flag-fill"),
  theme   = bslib::value_box_theme(bg = "#1d3461", fg = "#ffffff")
)
```

### Correlation Pattern

```{r}
bslib::value_box(
  title   = "Relationship type: diminishing returns at top",
  value   = "Logarithmic",
  showcase = bsicons::bs_icon("graph-up"),
  theme   = bslib::value_box_theme(bg = "#2d6a4f", fg = "#ffffff")
)
```

### Notable Exception

```{r}
bslib::value_box(
  title   = "High GDP but moderate life expectancy",
  value   = "Oil States",
  showcase = bsicons::bs_icon("exclamation-circle-fill"),
  theme   = bslib::value_box_theme(bg = "#c77d1a", fg = "#ffffff")
)
```

### Trend Direction

```{r}
bslib::value_box(
  title   = "Richer countries-longer lives, globally",
  value   = "Strong Positive",
  showcase = bsicons::bs_icon("arrow-up-right-circle-fill"),
  theme   = bslib::value_box_theme(bg = "#7b4f9e", fg = "#ffffff")
)
```

## Row {height="520px"}

### GDP per Capita vs Life Expectancy — Bubble Chart (2007)

```{r title = "GDP per Capita vs Life Expectancy: Bubble Chart (2007)"}
#| fig-height: 6
gap_2007_bubble <- gap_2007 %>%
  mutate(
    pop_size   = sqrt(pop) / 600,
    label_text = paste0(
      "<b>", country, "</b> (", region, ")<br>",
      "Life Exp: ", round(lifeExp, 1), " yrs<br>",
      "GDP/cap: $", comma(round(gdpPercap)), "<br>",
      "Population: ", comma(pop)
    )
  )

plot_ly(
  gap_2007_bubble,
  x         = ~gdpPercap,
  y         = ~lifeExp,
  color     = ~region,
  colors    = region_colors,
  size      = ~pop,
  sizes     = c(5, 60),
  type      = "scatter",
  mode      = "markers",
  text      = ~label_text,
  hoverinfo = "text",
  marker    = list(opacity = 0.72, sizemode = "diameter")
) %>%
  layout(
    title  = list(
      text = "Wealth vs. Health: GDP per capita vs. life expectancy (2007)",
      font = list(size = 14)
    ),
    xaxis  = list(
      title = "GDP per Capita (USD, log scale)",
      type  = "log",
      tickvals = c(500, 1000, 5000, 10000, 50000),
      ticktext = c("$500", "$1,000", "$5,000", "$10,000", "$50,000")
    ),
    yaxis  = list(title = "Life Expectancy (years)"),
    legend = list(orientation = "h", y = -0.15),
    margin = list(b = 60)
  )
```

## Row {height="440px"}

### Regional Averages Comparison (2007) {width="650"}

```{r title = "Regional Averages Comparison (2007)"}
#| fig-height: 5
region_indexed <- region_2007 %>%
  mutate(
    life_exp_norm = avg_lifeExp  / max(avg_lifeExp),
    gdp_norm      = avg_gdpPercap / max(avg_gdpPercap)
  ) %>%
  pivot_longer(cols = c(life_exp_norm, gdp_norm),
               names_to  = "indicator",
               values_to = "value") %>%
  mutate(indicator = recode(indicator,
    life_exp_norm = "Life Expectancy (indexed)",
    gdp_norm      = "GDP per Capita (indexed)"
  ))

plot_ly(
  region_indexed,
  x         = ~region,
  y         = ~value,
  color     = ~indicator,
  colors    = c(
    "Life Expectancy (indexed)" = "#2d6a4f",
    "GDP per Capita (indexed)"  = "#1d3461"
  ),
  type      = "bar",
  text      = ~paste0("<b>", region, "</b><br>", indicator, ": ", round(value, 2)),
  hoverinfo = "text"
) %>%
  layout(
    barmode = "group",
    title   = list(text = "Both indicators by region-indexed to max",
                   font = list(size = 13)),
    xaxis   = list(title = ""),
    yaxis   = list(title = "Indexed Value (% of max)",
                   tickformat = ".0%"),
    legend  = list(orientation = "h", y = 1.1),
    margin  = list(b = 40)
  )
```

### Correlated Trajectory Over Time (1952–2007) {width="650"}

```{r title = "Correlated Trajectory Over Time (1952–2007)"}
#| fig-height: 5
# Split into regular points and 2007 highlighted points
regions <- unique(region_year$region)

p12 <- plot_ly()

for (r in regions) {
  df_r <- region_year %>% filter(region == r) %>% arrange(year)

  # Path line with arrow via annotation — add line trace
  p12 <- p12 %>%
    add_trace(
      data      = df_r,
      x         = ~avg_gdpPercap,
      y         = ~avg_lifeExp,
      type      = "scatter",
      mode      = "lines+markers",
      name      = r,
      color     = I(region_colors[r]),
      text      = ~paste0(
                    "<b>", region, " — ", year, "</b><br>",
                    "GDP/cap: $", comma(round(avg_gdpPercap)), "<br>",
                    "Life Exp: ", round(avg_lifeExp, 1), " yrs"
                  ),
      hoverinfo = "text",
      marker    = list(size = 5),
      line      = list(width = 1.8),
      showlegend = TRUE
    ) %>%
    # Highlight 2007 point
    add_trace(
      data      = df_r %>% filter(year == 2007),
      x         = ~avg_gdpPercap,
      y         = ~avg_lifeExp,
      type      = "scatter",
      mode      = "markers",
      name      = r,
      color     = I(region_colors[r]),
      marker    = list(size = 10),
      hoverinfo = "skip",
      showlegend = FALSE
    )
}

p12 %>%
  layout(
    title  = list(text = "Trajectory: wealth vs. health (1952–2007)",
                  font = list(size = 13)),
    xaxis  = list(title = "GDP per Capita (USD 000s)",
                  tickformat = "$.1fk"),
    yaxis  = list(title = "Life Expectancy (years)"),
    legend = list(orientation = "h", y = -0.15),
    margin = list(b = 60)
  )
```

## Row {height="380px"}

### Notable Outliers — Countries That Defy the Wealth-Health Trend

```{r  title = "Countries That Defy the Wealth-Health Trend"}
DT::datatable(
  outliers %>%
    transmute(
      Country   = country,
      Region    = region,
      `GDP/cap (2007)` = dollar(round(gdpPercap)),
      `Life Exp (2007)` = paste0(round(lifeExp, 1), " yrs"),
      `Key Story` = Story
    ),
  rownames = FALSE,
  options  = list(
    dom        = "t",
    pageLength = 10,
    columnDefs = list(list(width = "40%", targets = 4))
  ),
  class = "compact stripe hover"
) %>%
  formatStyle(
    "Region",
    backgroundColor = styleEqual(
      names(region_colors), unname(region_colors)
    ),
    color = "white",
    fontWeight = "bold"
  )
```



# Data {data-icon="fa-table"}

## Row {height="60px"}

### {.no-title}

```{r}
htmltools::div(
  style = "padding: 6px 2px 0;",
  htmltools::h4(
    "Full Gapminder Dataset",
    style = "font-family:'Fraunces',serif; font-weight:500; margin:0 0 2px;"
  ),
  htmltools::p(
    "All 1,704 country-year observations from the gapminder package (1952–2007). 
     Use the search and filter controls to explore, then download the full or 
     filtered dataset as a CSV.",
    style = "font-size:13px; color:#6b6760; margin:0;"
  )
)
```


## Row {height="680px"}

### Full Dataset {.no-title}

```{r}
htmltools::browsable(
htmltools::tagList(

    reactable(
      gapminder %>%
        rename(
          Country   = country,
          Region    = continent,
          Year      = year,
          `Life Exp (yrs)` = lifeExp,
          Population       = pop,
          `GDP/cap ($)`    = gdpPercap
        ) %>%
        mutate(
          `GDP/cap ($)` = round(`GDP/cap ($)`, 2),
          `Life Exp (yrs)` = round(`Life Exp (yrs)`, 2)
        ),
      elementId           = "gapminder-table",
      searchable          = TRUE,
      filterable          = TRUE,
      striped             = TRUE,
      highlight           = TRUE,
      bordered            = TRUE,
      defaultPageSize     = 15,
      showPageSizeOptions = TRUE,
      pageSizeOptions     = c(10, 25, 50, 100),
      defaultSorted       = list(Year = "desc", `Life Exp (yrs)` = "desc"),
      theme = reactableTheme(
        headerStyle      = list(background = "#1d3461", color = "#ffffff",
                                fontWeight = 500, fontSize = 12),
        stripedColor     = "#f5f3ee",
        highlightColor   = "#eaf5ee",
        borderColor      = "rgba(0,0,0,0.07)",
        style            = list(fontFamily = "'DM Sans', sans-serif", fontSize = 13)
      ),
      columns = list(
        Region = colDef(
          maxWidth = 110,
          cell = function(value) {
            colors <- c(Africa="#2d6a4f", Americas="#1d3461", Asia="#c94f2a",
                        Europe="#7b4f9e", Oceania="#c77d1a")
            col <- colors[[as.character(value)]]
            htmltools::span(
              value,
              style = paste0(
                "background:", col, ";color:#fff;padding:2px 9px;",
                "border-radius:10px;font-size:11px;font-weight:500;"
              )
            )
          }
        ),
        Year = colDef(maxWidth = 70, align = "center"),
        Population = colDef(
          align  = "right",
          format = colFormat(separators = TRUE, digits = 0)
        ),
        `GDP/cap ($)` = colDef(
          align  = "right",
          format = colFormat(currency = "USD", separators = TRUE, digits = 2)
        ),
        `Life Exp (yrs)` = colDef(
          align    = "center",
          maxWidth = 130,
          cell = function(value) {
            bar_w <- round((value - 20) / 65 * 100)
            htmltools::div(
              style = "display:flex;align-items:center;gap:6px;",
              htmltools::div(
                style = paste0(
                  "width:", bar_w, "px;max-width:90px;height:7px;",
                  "border-radius:4px;background:#2d6a4f;"
                )
              ),
              htmltools::span(value)
            )
          }
        )
      )
    ),

    tags$br(),

    tags$div(
      style = "display:flex;gap:12px;flex-wrap:wrap;",

      tags$button(
        "\u2b07 Download full dataset as CSV",
        onclick = "Reactable.downloadDataCSV('gapminder-table', 'gapminder_full.csv')",
        style = paste0(
          "background-color:#1d3461;color:white;border:none;",
          "padding:9px 20px;border-radius:6px;cursor:pointer;",
          "font-size:14px;font-family:'DM Sans',sans-serif;"
        )
      ),

      tags$button(
        "\u2b07 Download filtered rows as CSV",
        onclick = "Reactable.downloadDataCSV('gapminder-table', 'gapminder_filtered.csv', {filtered: true})",
        style = paste0(
          "background-color:#2d6a4f;color:white;border:none;",
          "padding:9px 20px;border-radius:6px;cursor:pointer;",
          "font-size:14px;font-family:'DM Sans',sans-serif;"
        )
      )
    )

  )
)
```

# Discussion {data-icon="fa-comments"}

## Row {height="auto"}

### Overview & Objectives {width="650"}

```{r}
htmltools::div(
  style = "font-family:'DM Sans',sans-serif; line-height:1.75; font-size:14px; color:#1a1814;",

  htmltools::h4("What This Dashboard Explores",
    style = "font-family:'Fraunces',serif; font-weight:500; color:#1d3461; margin-bottom:10px;"),

  htmltools::p(
    "This dashboard analyses two core development indicators from the ", 
    htmltools::strong("Gapminder dataset"), " — ", 
    htmltools::em("life expectancy at birth"), " and ", 
    htmltools::em("GDP per capita (PPP-adjusted USD)"), 
    " — across 142 countries and 12 time periods spanning 1952 to 2007."
  ),
  htmltools::br(),
  htmltools::p(
    "The Gapminder data, popularised by Hans Rosling, captures the broad sweep of global 
     development over the second half of the twentieth century. By examining these two 
     indicators together, we can ask a deceptively simple question: ", 
    htmltools::em("does national wealth translate into longer lives?"), 
    " The short answer is yes — but with important nuances."
  ),
  htmltools::br(),
  htmltools::h4("Data Preparation",
    style = "font-family:'Fraunces',serif; font-weight:500; color:#1d3461; margin-bottom:10px;"),
  htmltools::p(
    "The ", htmltools::code("gapminder"), " R package provides a pre-cleaned, tidy dataset 
     requiring minimal manipulation. Key ", htmltools::code("dplyr"), " operations applied include:"
  ),
  htmltools::tags$ul(
    style = "padding-left:20px; margin-top:6px;",
    htmltools::tags$li(htmltools::code("filter()"), " — isolating single years (e.g. 2007, 1952) for cross-sectional analysis"),
    htmltools::tags$li(htmltools::code("group_by() + summarise()"), " — computing population-weighted regional averages per year"),
    htmltools::tags$li(htmltools::code("weighted.mean()"), " — ensuring large-population countries (China, India) do not bias simple averages"),
    htmltools::tags$li(htmltools::code("mutate()"), " — deriving new variables such as total GDP (pop × gdpPercap) and income brackets"),
    htmltools::tags$li(htmltools::code("pivot_longer() / pivot_wider()"), " — reshaping for grouped bar charts and growth comparisons"),
    htmltools::tags$li(htmltools::code("left_join()"), " — merging Gapminder records onto world sf geometry for choropleth mapping"),
    htmltools::tags$li(htmltools::code("slice_max() / slice_min()"), " — extracting top and bottom country rankings"),
    htmltools::tags$li(htmltools::code("reorder()"), " — sorting factor levels for ranked bar charts")
  )
)
```

### Key Findings — Life Expectancy {width="650"}

```{r}
htmltools::div(
  style = "font-family:'DM Sans',sans-serif; line-height:1.75; font-size:14px; color:#1a1814;",

  htmltools::h4("Life Expectancy: A World Transformed",
    style = "font-family:'Fraunces',serif; font-weight:500; color:#2d6a4f; margin-bottom:10px;"),

  htmltools::p(
    "Global average life expectancy rose from approximately ", htmltools::strong("49.9 years in 1952"),
    " to ", htmltools::strong("67.0 years in 2007"), " — a gain of over 17 years in just 55 years. 
     This is one of the most remarkable achievements in human history, driven by the 
     spread of vaccines, antibiotics, clean water infrastructure, and maternal health programmes."
  ),
  htmltools::br(),

  htmltools::div(
    style = "background:#eaf5ee;border-left:4px solid #2d6a4f;padding:12px 16px;border-radius:6px;margin-bottom:12px;",
    htmltools::strong("Asia's leap: "),
    "Asia recorded the sharpest regional improvement — from ~46 years in 1952 to ~71 years 
     by 2007. This reflects transformative public health investments in China, the post-war 
     reconstruction of Japan and South Korea, and India's gradual health system expansion."
  ),

  htmltools::div(
    style = "background:#fdf1ea;border-left:4px solid #c94f2a;padding:12px 16px;border-radius:6px;margin-bottom:12px;",
    htmltools::strong("Africa's stall: "),
    "Despite consistent improvement through the 1980s, Sub-Saharan Africa's life expectancy 
     plateaued and partially reversed through the 1990s and early 2000s, largely due to the 
     HIV/AIDS epidemic. Countries like Botswana, Swaziland, and Zimbabwe saw life expectancy 
     fall back below 1970 levels."
  ),

  htmltools::div(
    style = "background:#eef4fb;border-left:4px solid #1d3461;padding:12px 16px;border-radius:6px;",
    htmltools::strong("The North-South divide: "),
    "The world map reveals a persistent geographic pattern — Europe and East Asia cluster 
     above 75 years while most of Sub-Saharan Africa remains below 55 years, pointing to 
     deep structural inequalities in healthcare access, nutrition, and governance."
  )
)
```

## Row {height="auto"}

### Key Findings — GDP per Capita {width="650"}

```{r}
htmltools::div(
  style = "font-family:'DM Sans',sans-serif; line-height:1.75; font-size:14px; color:#1a1814;",

  htmltools::h4("GDP per Capita: Growth With Widening Gaps",
    style = "font-family:'Fraunces',serif; font-weight:500; color:#c77d1a; margin-bottom:10px;"),

  htmltools::p(
    "GDP per capita grew across all regions between 1952 and 2007, but at dramatically 
     different speeds. Europe and Oceania saw steady compound growth, while Asia's trajectory 
     accelerated sharply after 1980 — driven by the ", htmltools::em("East Asian economic miracle"),
    " in Japan, South Korea, Taiwan, and Singapore, and later by China's rapid industrialisation."
  ),
  htmltools::br(),

  htmltools::div(
    style = "background:#faeeda;border-left:4px solid #c77d1a;padding:12px 16px;border-radius:6px;margin-bottom:12px;",
    htmltools::strong("Extreme inequality: "),
    "By 2007, Norway's GDP per capita (~$49,000) was approximately ", 
    htmltools::strong("178 times"), " that of the Democratic Republic of Congo (~$277). 
     This ratio illustrates how global economic growth has been deeply uneven — 
     and how geography, institutions, and natural resources shape development outcomes."
  ),

  htmltools::div(
    style = "background:#eaf5ee;border-left:4px solid #2d6a4f;padding:12px 16px;border-radius:6px;margin-bottom:12px;",
    htmltools::strong("The income distribution: "),
    "The histogram of countries by income bracket reveals a bimodal pattern — a cluster 
     of low-income countries (mostly African) and a cluster of high-income countries 
     (Europe, Americas, Oceania), with a thinner middle. This 'missing middle' reflects 
     persistent structural traps that prevent low-income economies from graduating."
  ),

  htmltools::div(
    style = "background:#fdf1ea;border-left:4px solid #c94f2a;padding:12px 16px;border-radius:6px;",
    htmltools::strong("The oil exception: "),
    "Gulf states like Kuwait and Equatorial Guinea show that hydrocarbon wealth can 
     generate very high GDP per capita without equivalent broad-based development — 
     pointing to the well-documented 'resource curse' and its implications for 
     health and human capital investment."
  )
)
```

### Key Findings — The Relationship {width="650"}

```{r}
htmltools::div(
  style = "font-family:'DM Sans',sans-serif; line-height:1.75; font-size:14px; color:#1a1814;",

  htmltools::h4("Wealth and Health: A Logarithmic Story",
    style = "font-family:'Fraunces',serif; font-weight:500; color:#7b4f9e; margin-bottom:10px;"),

  htmltools::p(
    "The bubble chart on the Relationship page tells the central story of this dashboard: 
     richer countries, on average, enjoy longer lives. But the relationship is not linear 
     — it follows a ", htmltools::strong("logarithmic curve"), ". Doubling GDP per capita 
     from $500 to $1,000 is associated with a much larger gain in life expectancy than 
     doubling from $25,000 to $50,000."
  ),
  htmltools::br(),

  htmltools::div(
    style = "background:#eeedfe;border-left:4px solid #7b4f9e;padding:12px 16px;border-radius:6px;margin-bottom:12px;",
    htmltools::strong("Diminishing returns: "),
    "This diminishing-returns pattern suggests that at very low income levels, even modest 
     increases in national wealth — enough to fund basic healthcare, sanitation, and 
     nutrition — deliver dramatic health gains. Above roughly $15,000 per capita, 
     additional wealth produces progressively smaller life expectancy improvements."
  ),

  htmltools::div(
    style = "background:#eaf5ee;border-left:4px solid #2d6a4f;padding:12px 16px;border-radius:6px;margin-bottom:12px;",
    htmltools::strong("Notable overperformers: "),
    "Cuba and Sri Lanka stand out as countries that achieve life expectancies well above 
     what their GDP would predict — demonstrating that deliberate policy choices around 
     universal healthcare and education can partially decouple health from wealth."
  ),

  htmltools::div(
    style = "background:#fdf1ea;border-left:4px solid #c94f2a;padding:12px 16px;border-radius:6px;margin-bottom:12px;",
    htmltools::strong("Notable underperformers: "),
    "Botswana, Gabon, and Equatorial Guinea all have GDP per capita figures that would 
     predict much higher life expectancies than observed — highlighting how resource 
     extraction economies, disease burden (HIV/AIDS), and governance failures can 
     break the wealth-health link."
  ),

  htmltools::div(
    style = "background:#eef4fb;border-left:4px solid #1d3461;padding:12px 16px;border-radius:6px;",
    htmltools::strong("The trajectory over time: "),
    "The correlated trajectory chart shows that all regions moved toward both higher GDP 
     and higher life expectancy between 1952 and 2007 — but along very different paths 
     and at very different speeds, reinforcing that economic growth alone does not 
     guarantee human development. The data gives strong empirical support to the 
     view that both ", htmltools::em("economic growth"), " and ", 
    htmltools::em("targeted health investment"), " are necessary for sustained improvement."
  )
)
```

## Row {height="auto"}

### Conclusions & Limitations {.no-title}

```{r}
htmltools::div(
  style = paste0(
    "font-family:'DM Sans',sans-serif;line-height:1.75;font-size:14px;",
    "color:#1a1814;background:#f5f3ee;border-radius:10px;padding:20px 24px;"
  ),

  htmltools::h4("Conclusions",
    style = "font-family:'Fraunces',serif;font-weight:500;color:#1d3461;margin-bottom:10px;"),

  htmltools::p(
    "This dashboard demonstrates three broad conclusions. First, the world has made 
     extraordinary progress on life expectancy since 1952, driven by public health advances 
     rather than wealth alone. Second, GDP per capita has grown substantially but unequally, 
     with the gap between the richest and poorest countries widening in absolute terms even 
     as it has narrowed in relative terms in some regions. Third, wealth and health are 
     positively correlated in a logarithmic pattern, but the relationship has important 
     exceptions — both overperformers (Cuba, Sri Lanka) and underperformers (Gulf oil states, 
     HIV-affected Southern African nations) — reminding us that policy choices, disease 
     burden, and governance shape outcomes as much as income does."
  ),

  htmltools::br(),

  htmltools::h4("Limitations",
    style = "font-family:'Fraunces',serif;font-weight:500;color:#c94f2a;margin-bottom:10px;"),

  htmltools::tags$ul(
    style = "padding-left:20px;",
    htmltools::tags$li(
      htmltools::strong("Coverage gaps: "),
      "The Gapminder package covers 142 countries — many small island states, 
       conflict zones, and microstates are absent, which may skew regional averages."
    ),
    htmltools::tags$li(
      htmltools::strong("Temporal resolution: "),
      "Data is provided in 5-year intervals, smoothing over year-to-year shocks such 
       as famines, epidemics, and financial crises."
    ),
    htmltools::tags$li(
      htmltools::strong("No causal claims: "),
      "All relationships shown are descriptive and correlational. The dashboard does not 
       establish that higher GDP causes longer lives, or vice versa — both may be driven 
       by common underlying factors (institutions, education, geography)."
    ),
    htmltools::tags$li(
      htmltools::strong("Static endpoint: "),
      "The dataset ends in 2007. Significant events since then — the 2008 financial 
       crisis, the COVID-19 pandemic, and advances in global health — are not reflected."
    )
  )
)
```





